---
title: "Phase 2 - Week 1 Exercises"
author: "Tom Polak"
date: "October 2, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tibble)
library(ggplot2)
library(dplyr)
library(readr)
library(tidyverse)
library(hms)
```

##10.5 Exercises

**1. How can you tell if an object is a tibble? (Hint: try printing mtcars, which is a regular data frame).**

  A tibble will not show all columns nor all rows (limiting the # of rows displayed to 10). In this case when mtcars is printed, it shows all rows and columns in the console.

**2. Compare and contrast the following operations on a data.frame and equivalent tibble. What is different? Why might the default data frame behaviours cause you frustration?**

  
  `df <- data.frame(abc = 1, xyz = "a")`
  `df$x`
  `df[, "xyz"]`
  `df[, c("abc", "xyz")]`
  
  
  The first line will return all "variables" that have "x" as the leading letter. teh second line will return all variables called "xyz". The third line explicitly calls both variables ,"abc","xyz". It could be annoying to use the $x if there is another variable called "xerox" or "xylophone" because those would all be picked up as well. 
  

**3. If you have the name of a variable stored in an object, e.g. var <- "mpg", how can you extract the reference variable from a tibble?**

  You have to use [[]] "double brackets", if you just put in "var" then it would look for a variable called "var" within the tibble. 

**4. Practice referring to non-syntactic names in the following data frame by:**


```{r tibble}
annoying <- tibble(
  `1` = 1:10,
  `2` = `1` * 2 + rnorm(length(`1`))
)
```

i. Extracting the variable called 1
```{r example1}
annoying[["1"]]
```

ii. Plotting a scatterplot of 1 vs 2. 
``` {r example2}
ggplot(annoying, aes(x=`1`,y=`2`))+
  geom_point()
```

iii. Creating a new column called 3, which is 2 divided by 1
``` {r example3}
annoying[["3"]] <- annoying[["2"]] / annoying[["1"]]
annoying[["3"]]
```   


iv. Renaming the columns to one, two and three.
```{r example4}
annoying <- rename(annoying, one = `1`, two = `2`, three = `3`)
glimpse(annoying)
```

**5. What does tibble::enframe() do? When might you use it?**

  It converts named vectors to a data frame with names and values, kind of like a transpose function. 

**6. What option controls how many additional column names are printed at the footer of a tibble?**

`n_extra` is the option for the print.tbl_df function. 


## **11.2.2 Exercises**

**1. What function would you use to read a file where fields were separated with “|”?**

  You would use the "delim" function. For example: `read_delim(data, delim ="|")`

**2. Apart from file, skip, and comment, what other arguments do read_csv() and read_tsv() have in common?**

  They also have the following in common:
  * `col_names` & `col_types`
  * `locale`
  * `na` & `quoted_na control`
  * `trim_ws` 
  * `n_max`
  * `guess_max`
  * `progress`

**3. What are the most important arguments to read_fwf()?**

  I would argue the most important are where the columns start and end, aka `col_position`. It seems you always need to use another fwf function though with the `col_positions` such as a fwf_widths or fwf_positions. This could be tricky to use, but is fast. 

4. **Sometimes strings in a CSV file contain commas. To prevent them from causing problems they need to be surrounded by a quoting character, like " or '. By convention, read_csv() assumes that the quoting character will be ", and if you want to change it you’ll need to use read_delim() instead. What arguments do you need to specify to read the following text into a data frame?**

`"x,y\n1,'a,b'"`

```{r questions4}
x <- ("x,y\n1,'a,b'")
read_delim(x, ",", quote ="'")
```

**5. Identify what is wrong with each of the following inline CSV files. What happens when you run the code?**

`read_csv("a,b\n1,2,3\n4,5,6")`

In this first one, there are only TWO column headers, but three columns of data, so the third column is dropped (in the first row, "3" is dropped, in the second "6" is dropped)

`read_csv("a,b,c\n1,2\n1,2,3,4")`

In this one, there are not enough columns in the first row of data (not including headers), so it returns a null value for the third column. It also has TOO many columns in the next row of data, so it does not return the fourth column. 

`read_csv("a,b\n\"1")`
Skips the first row of data because there's nothing in there, and the second row only has one value. So column "b" has an NA.

`read_csv("a,b\n1,2\na,b")`

This one worked perfectly, and returned a nice 2x2 tibble

`read_csv("a;b\n1;3")`

This one created a 1x1 tibble with column header of "a;b" and one row of data with "1;3". If we wanted to use ";" as a parsing character we'd need to use read_delim.


## 11.3.5 Exercises

**1. What are the most important arguments to locale()?**

  `locale()` helps to navigate the many different ways countries put in numbers/dates in strings. I would argue, the most important arguments are `date_names`, `date_format`, `decimal_mark`, and `encoding`. 

**2. What happens if you try and set decimal_mark and grouping_mark to the same character? What happens to the default value of grouping_mark when you set decimal_mark to “,”? What happens to the default value of decimal_mark when you set the grouping_mark to “.”?**

Locale will throw an error when you try to set the character. 
If the decimal_mark is set to a comma then the grouping_mark will be set to a decimal (e.g. $120.000,98)
If the grouping_mark is set to a "," then the decimal_mark is set to a decimal (e.g. $120,000.98) 

**3. I didn’t discuss the date_format and time_format options to locale(). What do they do? Construct an example that shows when they might be useful.**

It returns default date time formats. This can be useful when different languages show dates differently in a data set. Example below:

```{r foreign}
parse_date("1 enero 2015", "%d %B %Y", locale = locale("es"))
```


**4. If you live outside the US, create a new locale object that encapsulates the settings for the types of file you read most commonly.**

If I live in Spanish America (I don't) then I could do this:

``` {r Spanish}
locale("es", decimal_mark = ",")
```

**5. What’s the difference between read_csv() and read_csv2()?**

  The delimiter, read_csv() uses a "," while read_csv2() uses ";"

**6. What are the most common encodings used in Europe? What are the most common encodings used in Asia? Do some googling to find out.**

  This website was helpful: https://stackoverflow.com/questions/8509339/what-is-the-most-common-encoding-of-each-language
  
  Although at the end of the day, I'll probably just google into it when it's necessary. 

**7. Generate the correct format string to parse each of the following dates and times:**

```{r question7}
d1 <- "January 1, 2010"
parse_date(d1, "%B %d, %Y")

d2 <- "2015-Mar-07"
parse_date(d2, "%Y-%b-%d")

d3 <- "06-Jun-2017"
parse_date(d3, "%d-%b-%Y")

d4 <- c("August 19 (2015)", "July 1 (2015)")
parse_date(d4, "%B %d %Y", c("(",")"))

d5 <- "12/30/14" # Dec 30, 2014
parse_date(d5, "%m/%d/%y")

t1 <- "1705"
parse_time(t1, "%H%M")

t2 <- "11:15:10.12 PM"
parse_time(t2, "%I:%M:%S %p")
```

