---
title: "Statistical Inference - CLT Simulation Project"
author: "Tom Polak"
date: "December 12, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)

```


## Overview
This project will investigate the exponential distribution in R and compare it with the Central Limit Theorem. For this analysis, a rate parameter of 0.2 will be set for all simulations. The analysis will then compare the distribution of averages of 40 exponentials over 1000 simulations. 


## Simulations

Set the lambda, exponentials, and number of simulations (n)

```{r Simulations SetUp Chunk, results= 'hide'}
lambda = 0.2
exponentials = 40
n = 1000
```

Run the simulation with the set variables

```{r Simulations Execute Chunk, results= 'hide'}
simMeans = NULL
for(i in 1:n) simMeans = c(simMeans, mean(rexp(exponentials,lambda)))
```

## Sample Mean vs Theoretical Mean

### Theoretical Mean

We know the theoretical mean is 1/lambda

```{r SampleMean - Theoretical Mean}
1/lambda
```

### Sample Mean

Calculating the sample mean from the simulations we ran earlier
```{r SampleMean - Sample Mean}
mean(simMeans)
```

### Comparison

```{r SampleMean - Compare the Sample and Theoretical}
abs(mean(simMeans) - 1/lambda)
```

There is only a slight difference between the theoretical mean and the sample mean created from the 1000 simulations of 40 exponentials. Next we'll take a look at the variance. 

## Sample Variance vs Theoretical Variance

### Theoretical Variance

The standard deviation of an exponential distribution is 1/lambda. Since the standard deviation is the square root of the variance, we can calculate the variance as (1/lambda)^2

```{r Variance - Theoretical Variance}
(1/(lambda*sqrt(exponentials)))^2
```

### Sample Variance

Calculating the sample variance from the simulations we ran earlier

```{r Variance - Sample Variance}
var(simMeans)
```

### Comparison

```{r Variance - Comparison}
abs(var(simMeans)-(1/(lambda*sqrt(exponentials)))^2)
```

There is only a slight difference between the theoretical variance and the sample variance.

## Distribution

Below is a density histogram plot of the 1000 simulations. The normal distribution (black curve) is overlaid on the simulation data (blue histograms).

``` {r Distribution - Plot}
ggplot(data.frame(simMeans), aes(x=simMeans)) +
  geom_histogram(aes(y=..density..), binwidth = 0.2, color = "black", fill = "light blue") +
  stat_function(fun=dnorm, args = list(mean = lambda^-1, sd=(lambda*sqrt(exponentials))^-1),size=2) +
  labs(
      title = "Plot of The Simulations", 
      x = "Simulation Mean",
      y = "Density"
  )
```


