---
title: "PA1_Template"
author: "Tom Polak"
date: "December 20, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading and preprocessing the data
##### 1. Load the data (i.e. read.csv())
```{r, results='markup', warning=FALSE, message=FALSE}

library(tidyverse)
library(ggplot2)

activityData <- read.csv('activity.csv')
```

##### 2. Process/transform the data (if necessary) into a format suitable for your analysis

-----

## What is mean total number of steps taken per day?
```{r}
stepsByDay <- activityData %>%
  na.omit() %>% 
  group_by(date) %>% 
    summarise(TotalSteps = sum(steps))
```

##### 1. Make a histogram of the total number of steps taken each day
```{r}
hist(stepsByDay$TotalSteps, 
     breaks=seq(from=0, to=25000, by=2500),
     col="red", 
     xlab="Total number of steps", 
     ylim=c(0, 30), 
     main="Histogram of the total number of steps taken each day")
```

##### 2. Calculate and report the mean and median total number of steps taken per day
```{r}
stepsByDayMean <- mean(stepsByDay$TotalSteps)
stepsByDayMedian <- median(stepsByDay$TotalSteps)
```
* Mean: `r stepsByDayMean`
* Median:  `r stepsByDayMedian`

-----

## What is the average daily activity pattern?
```{r}
averageStepsPerTimeBlock <- aggregate(x=list(meanSteps=activityData$steps), by=list(interval=activityData$interval), FUN=mean, na.rm=TRUE)
```

##### 1. Make a time series plot
```{r}
ggplot(data=averageStepsPerTimeBlock, aes(x=interval, y=meanSteps)) +
    geom_line() +
    xlab("5-minute interval") +
    ylab("average number of steps taken") 
```

##### 2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
mostSteps <- which.max(averageStepsPerTimeBlock$meanSteps)
timeMostSteps <-  gsub("([0-9]{1,2})([0-9]{2})", "\\1:\\2", averageStepsPerTimeBlock[mostSteps,'interval'])
```

* Most Steps at: `r timeMostSteps`

----

## Imputing missing values
##### 1. Calculate and report the total number of missing values in the dataset 
```{r}
numMissingValues <- length(which(is.na(activityData$steps)))
```

* Number of missing values: `r numMissingValues`

##### 2. Devise a strategy for filling in all of the missing values in the dataset.
##### 3. Create a new dataset that fills in the missing data
```{r}
activityDataImputed <- activityData

# Find the NA positions
na_pos <- which(is.na(activityData$steps))

# Create a vector of means
mean_vec <- rep(mean(activityData$steps, na.rm=TRUE), times=length(na_pos))

# Replace the NAs by the means
activityDataImputed[na_pos, "steps"] <- mean_vec
```


##### 4. Make a histogram of the total number of steps taken each day 
```{r}
# Compute the total number of steps each day (NA values removed)
sum_data <- aggregate(activityDataImputed$steps, by=list(activityDataImputed$date), FUN=sum)

# Rename the attributes
names(sum_data) <- c("date", "total")

# Compute the histogram of the total number of steps each day
hist(sum_data$total, 
     breaks=seq(from=0, to=25000, by=2500),
     col="blue", 
     xlab="Total number of steps", 
     ylim=c(0, 30), 
     main="Histogram of the total number of steps taken each day\n(NA replaced by mean value)")
```

##### ... and Calculate and report the mean and median total number of steps taken per day. 
```{r}
stepsByDayMeanImputed <- mean(sum_data$total)
stepsByDayMedianImputed <- median(sum_data$total)
```
* Mean (Imputed): `r stepsByDayMeanImputed`
* Median (Imputed):  `r stepsByDayMedianImputed`


----

## Are there differences in activity patterns between weekdays and weekends?
##### 1. Create a new variable in our new imputed data set that stores if the day is a weekend or weekday.

```{r}
activityDataImputed$dateType <-  ifelse(as.POSIXlt(activityDataImputed$date)$wday %in% c(0,6), 'weekend', 'weekday')
```

##### 2. Easiesty way to show this is through a panel plot.

```{r}
averagedActivityDataImputed <- aggregate(steps ~ interval + dateType, data=activityDataImputed, mean)
ggplot(averagedActivityDataImputed, aes(interval, steps)) + 
    geom_line() + 
    facet_grid(dateType ~ .) +
    xlab("5-minute interval") + 
    ylab("avarage number of steps")
```
