---
title: "Chapter 28 Exercises"
author: "Tom Polak"
date: "October 31, 2017"
output:
  html_document: default
---

```{r setup, include=FALSE}
library(tidyverse)
library(lubridate)

```

##28.2.1 Exercises

**1. Create one plot on the fuel economy data with customised title, subtitle, caption, x, y, and colour labels.**

```{r chunk - 28.2.1-Q1}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(se = FALSE) +
  labs(
        title = "Larger Engines are BAD for the Environment",
       subtitle = "A subset of data PROVING LARGE CARS are killing the environment",
       caption = "The title and subtitle is a joke",
       x = "Size of Engine (ltrs)", 
       y = "Highway Miles per Gallon (mpg)",
       color = "Car Type"
       )
```


**2. The geom_smooth() is somewhat misleading because the hwy for large engines is skewed upwards due to the inclusion of lightweight sports cars with big engines. Use your modelling tools to fit and display a better model.**

```{r chunk - 28.2.1-Q2, echo=FALSE}
ggplot(mpg, aes(displ, hwy)) +
  geom_point(aes(color = class)) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    title = "Larger Engines are BAD for the Environment",
    subtitle = "A subset of data PROVING LARGE CARS are killing the environment",
    caption = "The title and subtitle is a joke",
    x = "Size of Engine (ltrs)", 
    y = "Highway Miles per Gallon (mpg)",
    color = "Car Type"
  )
```


**3. Take an exploratory graphic that you've created in the last month, and add informative titles to make it easier for others to understand.**

```{r chunk - 28.2.1-Q3, echo=FALSE}
who_clean <- who %>%
  gather(code, value, new_sp_m014:newrel_f65, na.rm = TRUE) %>% 
  mutate(code = stringr::str_replace(code, "newrel", "new_rel")) %>%
  separate(code, c("new", "var", "sexage")) %>% 
  select(-new, -iso2, -iso3) %>% 
  separate(sexage, c("sex", "age"), sep = 1)

whoSummary <- who_clean %>%
  group_by(country, year, sex) %>%
  filter(year > 1995) %>%
  summarise(cases = sum(value)) %>%
  unite(country_sex, country, sex, remove = FALSE) 

ggplot(whoSummary,aes(x = year, y = cases, group = country_sex, colour = sex)) +
  geom_line() +
  theme(legend.position = "bottom") +
  labs(
        title = "Cases of TB in each country by Year",
        captoin = "Graph originally from Chapter 12.6.1 R Exercise Textbook HW",
        x = "Year (1980-2013)",
        y = "TB Cases",
        color = "Sex"
      )
```


##28.3.1 Exercises

**1. Use geom_text() with infinite positions to place text at the four corners of the plot.**

```{r chunk - 28.3.1-Q1, echo=FALSE}
label <- tribble(
  ~displ, ~hwy, ~label, ~vjust, ~hjust,
  -Inf,  Inf,    "Y", "top", "left",
  Inf,  Inf,    "M", "top", "right",
  Inf, -Inf,    "C", "bottom", "right",
  -Inf, -Inf,    "A", "bottom", "left"
)

ggplot(mpg, aes(displ,hwy)) +
  geom_point() +
  geom_text(aes(label = label, vjust=vjust, hjust=hjust ), data = label )
```

**2. Read the documentation for annotate(). How can you use it to add a text label to a plot without having to create a tibble?**

You can pass the aesthetic mappings directly as arguments.

```{r chunk - 28.3.1-Q2}
ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  annotate("text", x = Inf, y = Inf, 
           label = "Increasing engine size is \nrelated to decreasing fuel economy.", vjust = "top", hjust = "right")
```

**3. How do labels with geom_text() interact with faceting? How can you add a label to a single facet? How can you put a different label in each facet? (Hint: think about the underlying data.)**

If you don't specify, then the label is added to all faceted graphs:

```{r chunk - 28.3.1-Q3A}
label <- tibble(
  displ = Inf,
  hwy = Inf,
  label = "Increasing engine size is \nrelated to decreasing fuel economy."
)

ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  geom_text(aes(label = label), data = label, vjust = "top", hjust = "right",
            size = 2) +
  facet_wrap(~ class)
```


If you just wanted to add to a single facet, then in your "label" tribble that you create, include a column with whichever variable you want the facet included in:

```{r chunk - 28.3.1-Q3B}
label <- tibble(
  displ = Inf,
  hwy = Inf,
  class = "2seater",
  label = "Increasing engine size is \nrelated to decreasing fuel economy."
)

ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  geom_text(aes(label = label), data = label, vjust = "top", hjust = "right",
            size = 2) +
  facet_wrap(~ class)
```

If you want a unique label for each graph, then you can do that as well in the tibble:

```{r chunk - 28.3.1-Q3C}
label <- tibble(
  displ = Inf,
  hwy = Inf,
  class = unique(mpg$class),
  label = stringr::str_c("Label for ", class)
)

ggplot(mpg, aes(displ, hwy)) +
  geom_point() +
  geom_text(aes(label = label), data = label, vjust = "top", hjust = "right",
            size = 3) +
  facet_wrap(~ class)
```

**4. What arguments to geom_label() control the appearance of the background box?**

* `label.padding`: padding around label
* `label.r`: amount of rounding in the corners
* `label.size`: size of label border

**5. What are the four arguments to arrow()? How do they work? Create a series of plots that demonstrate the most important options.**

The four arguments are:

* `angle`: Describes the width of the arrow head
* `length`: Length of the arrow (from tip to base)
* `ends`: One of "last","first", or "both" indicating which part of the line to draw the arrow heads
* `type`: Either "open" or "closed" indicating whether the arrowhead is filled or not

##28.4.4 Exercises

**1. Why doesn't the following code override the default scale?**
```{r chunk - 28.4.4-Q1A, eval = FALSE}
ggplot(df, aes(x, y)) +
  geom_hex() +
  scale_colour_gradient(low = "white", high = "red") +
  coord_fixed()

```

You need to use the scale_fill_gradient. According to the ggplot cheatsheet, geom_hex() uses a fill instead of a color argument. Try this instead:

```{r chunk - 28.4.4-Q1B}
df <- tibble(
  x = rnorm(10000),
  y = rnorm(10000)
)

ggplot(df, aes(x, y)) +
  geom_hex() +
  scale_fill_gradient(low = "white", high = "red") +
  coord_fixed()
```


**2. What is the first argument to every scale? How does it compare to labs()?**

The first argument is the label, so it is equivalent to using the labs function.

**3. Change the display of the presidential terms by:**

*1. Combining the two variants shown above.*
*2. Improving the display of the y axis.*
*3. Labelling each term with the name of the president.*
*4. Adding informative plot labels.*
*5. Placing breaks every 4 years (this is trickier than it seems!).*

```{r chunk - 28.4.4-Q3}
#Create the TermBreak vector to store the date range between the first year in dataset and last year, with four #years as the break point.
TermBreak <- make_date(
    seq(
        year(min(presidential$start)),
        year(max(presidential$end)),
        by = 4
        ),1,1)

presidential %>%
  mutate(id = 33 + row_number(), 
         name_id = stringr::str_c(name, " (", id, ")")
         ) %>%
  
  ggplot(aes(start, id, colour = party)) +
    geom_point() +
    geom_segment(aes(xend = end, yend = id)) +
    geom_label(aes(label = name_id),nudge_y = .5, nudge_x = 2, alpha=0.5) +
    
    scale_colour_manual(values = c(Republican = "red", Democratic = "blue")) +
    scale_y_continuous("President #",breaks = seq(33, 45, by = 1)) +
    scale_x_date(NULL, breaks = TermBreak, date_labels = "'%y") 
```


**4.Use override.aes to make the legend on the following plot easier to see.**

```{r chunk - 28.4.4-Q4A, eval=FALSE}
ggplot(diamonds, aes(carat, price)) +
  geom_point(aes(colour = cut), alpha = 1/20)
```

Cleaned up the above chunk in these three ways:

* Shifted the legend to the bottom
* Put the legend in one row
* Changed the alpha to 1 for the legend (instead of 1/20 for the plotted points) 

```{r chunk - 28.4.4-Q4B}
ggplot(diamonds, aes(carat, price)) +
  geom_point(aes(colour = cut), alpha = 1/20) +
  theme(legend.position = "bottom") +
  guides(colour = guide_legend(nrow = 1, override.aes = list(alpha = 1))) 
```



